\section{Evaluation of DYNAMOS Streaming Implementations}
\label{sec:dynamos-rq3}

This section addresses RQ3 by validating the implemented DYNAMOS data paths on a large-result workload. RQ2 explains how the variants are integrated. RQ3 measures whether those variants improve completion time, time-to-first-result, and resource usage compared with the classic unary baseline.

\subsection{Benchmark method}

The primary benchmark uses the \texttt{dataThroughTtp} archetype with a single data provider, \texttt{UVA}. The request still follows the TTP/SURF compute-provider path defined by the archetype, but the data fanout is intentionally limited to one provider so that the experiment isolates large-result transfer rather than multi-provider coordination.

Each request executes a bulk projection query over the large \texttt{PersonenLarge} and \texttt{AanstellingenLarge} tables and joins them on \texttt{Unieknr}. The result is limited to 50{,}000, 100{,}000, or 250{,}000 rows. The chunked variants use 5{,}000 rows per SQL chunk, which gives 10, 20, and 50 total chunks respectively. This value is an experimental configuration rather than a claimed optimum: it is large enough to avoid turning the benchmark into a per-row messaging test and small enough to expose partial progress. In the saved 50k-row bulk output, one 5{,}000-row chunk serialized as a DYNAMOS protobuf \texttt{MicroserviceCommunication} is approximately 280{,}255 bytes, or 273.7~KiB.

Four implementations are compared. Classic unary is taken from the original DYNAMOS \texttt{main} branch at commit \texttt{a2d9531}. The chunked unary, intra-job gRPC streaming, and RabbitMQ Streams variants are taken from the streaming implementation branch at commit \texttt{923b018}. Every primary \texttt{dataThroughTtp} cell uses 20 warm repetitions. Runs are accepted only when row counts match the expected limit and canonical content hashes validate the returned content. For chunked variants, the benchmark also requires partial result events before completion.

Two focused scenario checks are added after the primary benchmark. The first changes the archetype to \texttt{computeToData}. The second keeps \texttt{dataThroughTtp} but uses two data providers, \texttt{UVA,VU}. These checks test whether chunk preservation remains compatible when the generated job topology and aggregation path differ from the primary single-provider TTP case.

Resource usage is sampled from Kubernetes once per second across the relevant DYNAMOS namespaces and summarized as peak total CPU millicores and peak total memory in MiB. These sampled peaks are useful for comparing memory pressure, but they may underestimate short instantaneous spikes between samples. The measured request window starts immediately before the benchmark client submits the HTTP request and ends when the final response or final NDJSON event is observed. This means policy handling, orchestration, generated job creation, pod startup, data transfer, and response serialization are inside the measured window for both classic and chunked variants. The classic unary baseline required generated \texttt{surf}, \texttt{uva}, and \texttt{vu} jobs to be cleaned between repetitions. That cleanup happens before and after a measured request, not inside the request timing. Without it, later repetitions could return fast 502 or empty-provider responses because stale generated jobs were still present. The cleaned baseline is therefore the comparison used here.

\subsection{Completion latency}

\begin{table}[!t]
    \centering
    \scriptsize
    \caption{Average completion latency for DYNAMOS bulk retrieval. All cells completed 20/20 valid repetitions.}
    \label{tab:dynamos-rq3-latency}
    \setlength{\tabcolsep}{3pt}
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}{lrrr}
        \toprule
        Implementation & 50k rows & 100k rows & 250k rows \\
        \midrule
        Classic unary & 13.988s & 19.343s & 29.078s \\
        Chunked unary & 6.536s & 6.754s & 11.684s \\
        Intra-job gRPC streaming & 6.346s & 8.255s & 12.319s \\
        RabbitMQ Streams boundary & 5.948s & 7.727s & 14.066s \\
        \bottomrule
    \end{tabular}}
\end{table}

\begin{figure*}[!t]
    \centering
    \begin{tikzpicture}
        \begin{axis}[
            width=0.86\textwidth,
            height=5.2cm,
            xlabel={Rows returned},
            ylabel={Average completion time (s)},
            xtick={50000,100000,250000},
            xticklabels={50k,100k,250k},
            scaled x ticks=false,
            ymin=0,
            grid=major,
            legend columns=2,
            legend style={at={(0.5,-0.23)},anchor=north,draw=none},
            tick label style={font=\scriptsize},
            label style={font=\scriptsize},
            legend style={font=\scriptsize}
        ]
            \addplot+[mark=*] coordinates {(50000,13.988) (100000,19.343) (250000,29.078)};
            \addlegendentry{Classic unary}
            \addplot+[mark=square*] coordinates {(50000,6.536) (100000,6.754) (250000,11.684)};
            \addlegendentry{Chunked unary}
            \addplot+[mark=triangle*] coordinates {(50000,6.346) (100000,8.255) (250000,12.319)};
            \addlegendentry{Intra-job gRPC streaming}
            \addplot+[mark=diamond*] coordinates {(50000,5.948) (100000,7.727) (250000,14.066)};
            \addlegendentry{RabbitMQ Streams boundary}
        \end{axis}
    \end{tikzpicture}
    \caption{Average completion time for the local DYNAMOS bulk retrieval benchmark. Chunking is the dominant improvement: all chunk-aware variants scale better than classic unary as the returned row count grows.}
    \label{fig:dynamos-rq3-latency}
\end{figure*}

Table~\ref{tab:dynamos-rq3-latency} and Figure~\ref{fig:dynamos-rq3-latency} show that all three chunk-aware variants reduce average completion latency compared with classic unary. At 250k rows, chunked unary is 2.49x faster than classic unary, intra-job gRPC streaming is 2.36x faster, and the RabbitMQ Streams boundary variant is 2.07x faster. The 100k row case is especially favourable to chunked unary, which finishes in 6.754s compared with 19.343s for classic unary. The relatively small increase of the chunked variants between 50k and 100k rows is explained by the fixed control-path cost analysed with Table~\ref{tab:dynamos-rq3-first}: at these sizes much of the measured time is policy handling, orchestration, and job startup rather than transfer.

The average values include the long-tail runs that occur in the local Kubernetes deployment, so they are more conservative than the medians. At 250k rows, classic unary ranged from 21.901s to 50.854s. Chunked unary ranged from 7.544s to 23.233s, intra-job gRPC streaming from 7.490s to 23.311s, and RabbitMQ Streams from 9.968s to 26.520s. This spread shows why repeated runs are necessary. The source of this large variance is due to Kubernetes and DYNAMOS irregularities, not the actual transport layer.

\subsection{Additional scenario checks}

\begin{table*}[!t]
    \centering
    \scriptsize
    \caption{Focused DYNAMOS scenario checks beyond the primary single-provider \texttt{dataThroughTtp} benchmark.}
    \label{tab:dynamos-rq3-scenarios}
    \setlength{\tabcolsep}{3pt}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{llrrrrrr}
        \toprule
        Scenario & Providers & Rows per provider & Total rows & Classic unary & Chunked unary & Intra-job gRPC streaming & RabbitMQ Streams boundary \\
        \midrule
        \texttt{computeToData} & \texttt{UVA,VU} & 50k & 100k & 9.622s & 5.178s & 4.528s & 4.586s \\
        \texttt{computeToData} & \texttt{UVA,VU} & 250k & 500k & 16.620s & 9.182s & 7.989s & 10.289s \\
        \texttt{dataThroughTtp} & \texttt{UVA,VU} & 50k & 100k & 13.750s & 5.196s & 4.474s & 6.492s \\
        \texttt{dataThroughTtp} & \texttt{UVA,VU} & 250k & 500k & 35.396s & 12.032s & 11.165s & 15.618s \\
        \bottomrule
    \end{tabular}}
\end{table*}

Table~\ref{tab:dynamos-rq3-scenarios} shows that the chunk-aware paths are not limited to the primary single-provider TTP case. For \texttt{computeToData}, the improvement is smaller because the request avoids part of the TTP aggregation path, but the chunked variants still complete faster than classic unary at both tested sizes. For multi-provider \texttt{dataThroughTtp}, the improvement remains large: at 250k rows per provider, the classic baseline returns 500k total rows in 35.396s, while the chunked variants complete in 11.165s to 15.618s.

These scenario checks are compatibility-oriented rather than an exhaustive benchmark matrix. Their value is that they show the implementation still works when the archetype and provider count change, and that the speedup is strongest when DYNAMOS must move large row sets through the TTP path.

\subsection{Time to first result}

\begin{table}[!t]
    \centering
    \scriptsize
    \caption{Time-to-first-result and partial result behaviour. Classic unary has no partial results, so first result equals completion.}
    \label{tab:dynamos-rq3-first}
    \setlength{\tabcolsep}{3pt}
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}{lrrrrr}
        \toprule
        Implementation & Rows & First result & Done & Post-first & Partial chunks \\
        \midrule
        Classic unary & 50k & 13.988s & 13.988s & 0.000s & 0 \\
        Classic unary & 100k & 19.343s & 19.343s & 0.000s & 0 \\
        Classic unary & 250k & 29.078s & 29.078s & 0.000s & 0 \\
        Chunked unary & 50k & 5.778s & 6.536s & 0.758s & 9 \\
        Chunked unary & 100k & 4.977s & 6.754s & 1.777s & 19 \\
        Chunked unary & 250k & 5.145s & 11.684s & 6.539s & 49 \\
        Intra-job gRPC streaming & 50k & 5.574s & 6.346s & 0.772s & 9 \\
        Intra-job gRPC streaming & 100k & 6.250s & 8.255s & 2.005s & 19 \\
        Intra-job gRPC streaming & 250k & 6.017s & 12.319s & 6.302s & 49 \\
        RabbitMQ Streams boundary & 50k & 5.184s & 5.948s & 0.763s & 9 \\
        RabbitMQ Streams boundary & 100k & 5.416s & 7.727s & 2.311s & 19 \\
        RabbitMQ Streams boundary & 250k & 5.613s & 14.066s & 8.453s & 49 \\
        \bottomrule
    \end{tabular}}
\end{table}

Table~\ref{tab:dynamos-rq3-first} shows the qualitative benefit of chunking. Classic unary cannot expose progress: the first result is the final result. The chunked variants emit partial results substantially earlier. At 250k rows, the first partial result appears after about 5.1s to 6.0s on average, while completion follows after 11.7s to 14.1s. The post-first column estimates how much of the measured time remains after the first visible chunk. It is not a pure network-transfer measurement, because the first chunk still includes policy handling, job creation, and initial SQL execution. It is nevertheless useful because it separates request startup from the observable chunk-drain phase. Read together with Table~\ref{tab:dynamos-rq3-latency}, this decomposition shows that the end-to-end ratios understate the data-path difference: every chunked variant spends roughly 5--6~s before the first chunk, which is the shared policy, orchestration, and job-startup cost. End-to-end remains the headline metric since it is what a DYNAMOS client experiences, but the time-to-first-result values show that chunking also changes the interaction pattern: a client can observe progress and begin processing data before the full result has been materialized. This decomposition also matters for diagnosing whether a large request is still making progress.

\subsection{Resource usage}

\begin{table}[!t]
    \centering
    \scriptsize
    \caption{Peak sampled resource usage during DYNAMOS bulk retrieval. CPU is peak total millicores; memory is peak total MiB across sampled Kubernetes namespaces.}
    \label{tab:dynamos-rq3-resources}
    \setlength{\tabcolsep}{3pt}
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}{lrrr}
        \toprule
        Implementation & Rows & Peak CPU & Peak memory \\
        \midrule
        Classic unary & 50k & 958m & 1{,}571 MiB \\
        Classic unary & 100k & 813m & 2{,}046 MiB \\
        Classic unary & 250k & 1{,}198m & 3{,}695 MiB \\
        Chunked unary & 50k & 1{,}059m & 1{,}151 MiB \\
        Chunked unary & 100k & 961m & 1{,}723 MiB \\
        Chunked unary & 250k & 1{,}097m & 1{,}378 MiB \\
        Intra-job gRPC streaming & 50k & 1{,}180m & 1{,}198 MiB \\
        Intra-job gRPC streaming & 100k & 1{,}154m & 1{,}230 MiB \\
        Intra-job gRPC streaming & 250k & 1{,}418m & 1{,}495 MiB \\
        RabbitMQ Streams boundary & 50k & 684m & 1{,}161 MiB \\
        RabbitMQ Streams boundary & 100k & 982m & 1{,}436 MiB \\
        RabbitMQ Streams boundary & 250k & 1{,}687m & 1{,}715 MiB \\
        \bottomrule
    \end{tabular}}
\end{table}

The resource data in Table~\ref{tab:dynamos-rq3-resources} support the main architectural claim. At 250k rows, classic unary reaches 3{,}695 MiB peak memory. Chunked unary stays at 1{,}378 MiB, intra-job gRPC streaming at 1{,}495 MiB, and the RabbitMQ Streams boundary variant at 1{,}715 MiB. Chunking therefore reduces memory pressure consistently.

CPU is more nuanced. Chunked unary uses slightly less peak CPU than classic unary at 250k rows, but intra-job gRPC streaming and the RabbitMQ Streams boundary variant peak higher. This is expected. Chunking replaces one large buffer with more messages, more metadata handling, more serialization boundaries, and, in the RabbitMQ Streams variant, broker stream handling and reassembly. The defensible claim is therefore that chunking improves latency and memory. CPU remains transport-dependent.

\subsection{Failure modes and robustness}

During the benchmark work, several failure modes had to be controlled before the results could be interpreted reliably. Appendix~\ref{app:dynamos-failure-modes} summarizes these cases. The important distinction is that some fixes make the baseline valid, such as correcting the multi-provider merge, while others show the architectural benefit of chunking, such as avoiding single-message size limits and full-result buffering. The RQ3 robustness claim is therefore intentionally bounded: chunking removes size-dependent failure modes, reduces full-result memory pressure, and exposes partial progress, but it does not yet prove recovery from arbitrary pod crashes, broker outages, or mid-stream policy revocation inside DYNAMOS.

\subsection{Discussion and RQ3 answer}

The DYNAMOS-specific benchmark changes the interpretation of the isolated RQ1 transport study. RQ1 showed that client-streaming gRPC is the strongest raw transport in the controlled pipeline. The DYNAMOS evaluation shows that integration fit matters as much as raw transport speed. DYNAMOS is a policy-driven, dynamically composed middleware in which decoupling, generated job lifecycle, sidecar communication, and diagnosable failure behaviour are part of the design problem.

The central RQ1 hypothesis partly survives validation, but in a more precise form. Chunking is the main source of the improvement, not the name of the transport alone. RabbitMQ Streams remains useful to study as a distributed-boundary mechanism, but in the local DYNAMOS runs it adds overhead compared with ordinary chunked transfer. This is why the RQ3 conclusion ranks chunked unary highest for the current local system while keeping RabbitMQ Streams as a future distributed candidate.

For the current local DYNAMOS implementation, chunked unary is the strongest default candidate. It has the best average completion time at 100k and 250k rows, the lowest 250k memory peak, and the smallest architectural change from the original message-driven design. It also shows why the main bottleneck was not simply "gRPC versus broker", but full-result buffering across the DYNAMOS path.

Intra-job gRPC streaming remains a strong technical option. It is close to chunked unary at 250k rows and slightly faster at 50k rows. Its main weakness is architectural. Long-lived direct streams couple producer and consumer lifetime more tightly, which complicates dynamic composition, restart handling, and load balancing in Kubernetes-style deployments \cite{starck2023streaminglb}. It is therefore best interpreted as a targeted optimisation for controlled in-pod or direct service-to-service flows.

The RabbitMQ Streams boundary variant should be interpreted carefully. It beats classic unary because it uses the common chunk-aware path, not because RabbitMQ Streams is faster than normal RabbitMQ chunks in the local benchmark. Compared with chunked unary, it is slower and uses more peak CPU and memory in this run. Its value is therefore not local latency. Its value is the stream-specific broker model: retained records, offset-based replay, non-destructive consumption, and a closer fit for distributed inter-agent transfer.

The answer to RQ3 is that the implemented chunk-aware DYNAMOS data paths improve local bulk-retrieval scalability compared with classic unary. They reduce completion time, expose earlier partial results, and lower peak memory usage for 50k, 100k, and 250k row result sets. The best current local default is chunked unary. Intra-job gRPC streaming is a close alternative for pod-local service chains. RabbitMQ Streams should be kept as a distributed-boundary candidate, but its advantage must be validated under cross-site conditions rather than inferred from the local latency result.

\subsection{Threats to validity}

The RQ3 evidence has five main limits. First, the benchmark runs on a local Kubernetes deployment, so it does not capture cross-site latency, bandwidth variation, or independent administrative domains. Second, the additional scenario checks cover two alternative request shapes, not the full space of DYNAMOS archetypes, query forms, and provider combinations. They support compatibility, but they do not replace a broader workload campaign. Third, resource usage is sampled once per second, so peak CPU and memory values can miss short spikes. Fourth, the fixed 5{,}000-row chunk size is justified experimentally but not optimized across all workloads. Fifth, the measured request window includes policy handling, orchestration, generated job startup, and response serialization. That is fair for comparing user-visible DYNAMOS behaviour, but it compresses the apparent transfer-only speedup because fixed control-path costs remain present in every implementation.
